Multi-Domain Adaptation in Neural Machine Translation with Dynamic Sampling Strategies

### 2021-08-04 2021/wmt/auto_cl_mdmt.tex
quick review: 8 pages

\begin{document}
\maketitle
\setlength{\abovedisplayskip}{2pt}
\setlength{\belowdisplayskip}{2pt}
\begin{abstract}
  Building effective Neural Machine Translation models often implies accommodating diverse sets of heterogeneous data so as to optimize performance for the domain(s) of interest. Such multi-source / multi-domain adaptation problems are typically approached through instance selection or reweighting strategies, based on a static assessment of the relevance of training instances with respect to the task at hand. In this paper, we study dynamic data selection strategies that are able to automatically re-evaluate the usefulness of data samples and to evolve a data selection policy in the course of training. Based on the results of multiple experiments, we show that such methods constitute a generic framework to automatically and effectively cater to a variety of real-world situations, from multi-source domain adaptation to multi-domain learning and unsupervised domain adaptation. 
\end{abstract}

@@

It is usual in MT to opportunistically collect training samples from several domains, which means that training instances are distributed according to a mixture $\mathcal{D}^s$ such that $\mathcal{D}^s(x) = \sum_{d=1}^{n_d} \lambda^{s}(d) \mathcal{D}_d(x)$, with $\{\lambda^{s}(d), d=1 \dots n_d\}$ the mixture weights satisfying $\sum_d \lambda^{s}(d)=1$. To clarify the ambiguity in the notation of $\lambda$ and $\vlambda$, regardless the superscripts and the subscripts, $\lambda$ is used to denote a scalar and $\vlambda$ is used to denote a vector, for example, $\lambda^{t}(d)$ is the $d^{th}$ element of the vector $\vlambda^{t}$, which represents the probability distribution over the set of domains.

The main challenge is then to make the best of this heterogeneous data, with the aim to achieve the optimal performance for the intended test conditions. These might correspond to data from just one of the training domains, as in standard supervised (multi-source) domain adaptation; a more difficult case is when the test data is from one domain unseen in training (unseen domain adaptation); in multi-domain adaptation finally, the test distribution is itself a mixture of domains, some of which may also be observed in training.  Without loss of generality, one may then assume that the test distribution takes the form $\mathcal{D}^{t}(x) = \sum_d \lambda^{t}(d) \mathcal{D}_d(x)$ - with only one non-null  component in the case of pure domain adaptation. These settings are illustrated on Figure~\ref{fig:mdmt-lambdas}.
\begin{figure}[h]
  \centering
  \includegraphics[width=0.48\textwidth]{mdmt-lambdas}
  \caption{Training and testing with distribution mismatch. We consider just three domains, and represent vectors of mixture weights $\vlambda^{s}$ and $\vlambda^{t}$ in the 3-dimensional simplex. Training with weights in (a) and testing with weights in (c) is supervised multi-source domain adaptation to domain~2 ($d_2$), while (b)-(c) is the unsupervised version, with no training data from $d_2$; training with weights in (a) and testing with weights in (d) is multi-domain learning, also illustrated with configurations (a)-(e) (training domain $d_1$ is not seen in test), and (b)-(d)  (test domain $d_2$ is unseen in training).}\label{fig:mdmt-lambdas}
\end{figure}

@@

A widely used approach to supervised DA is \emph{fine-tuning} \citep{Luong15stanford,Freitag16fast}, where $\vlambda^{l}$ is allowed to vary during learning. With our notations, this approach amounts to first learning an initial parameter value with all the data ($\forall d, \lambda^{l}(d) = \lambda^{s}(d)$), then to continue training with only batches from the test domain $d_t$ ($\lambda^{l}(d) = \indic{d = d_t}$) with $\indic{A}$ the indicator function for predicate $A$. Note that this strategy is potentially suboptimal, as some out-of-domain samples may contribute to the final performance due to eg.\ domain overlap. Optimizing the learning distribution in multi-domain settings is even more challenging, as the learner has to best take advantage of potential domains overlaps, and also of the fact that some domains might be easier to learn than others.\fyDone{How to measure this?}
% \revision{Answer: The graphic of \system{Mixed-0} shows that with the same amount of data(batches) the convergences of the domains vary significantly}

@@

% Our method
MDAC constructs an adaptative training distribution $\vlambda^{l}$ that optimizes the data selection policy along with the training of the NMT model. We parameterize $\vlambda^{l}$ by a differentiable function $\vlambda^l(\psi)$, which will be described in detail in Section \ref{ssec:dds-sys}. We divide the training into many short sessions; in each session $t$, the model is trained with a static data distribution $\vlambda^{l}(\psi_t)$. After one learning session, we update the data distribution using the REINFORCE algorithm of \citet{Williams92simple}. The evolution of $\psi$ is thus defined by:
\begin{align*}
\psi_{t+1} &= \psi_t + \mathbf{lr}_{1} * \displaystyle{\mathop{\sum}_{d=1}^{n_d}} R(d) * \frac{\partial \lambda^l(d;\psi_t)}{\partial \psi}, \\
\end{align*}
\begingroup
\allowdisplaybreaks
where reward $R(d)$ is computed as:
\begin{align*}
  R(d) = J^t(\theta_{t+k},\vlambda^t) - J^t(\theta_t,\vlambda^t),
\end{align*}
and where we also define:
\begin{equation}
\begin{array}{rcl}
\theta_{t+i} &=& Update\big(\theta_{t+i-1},[x^i_j,y^i_j]_{j=1}^N\big) \\ \nonumber
x^i_j, y^i_j &\sim& \widehat{D^l_d}(x) \\
J^t(\theta,\vlambda^t) &=& \displaystyle{\mathop{\sum}_{d=1}^{n_d}}\lambda^t(d)\displaystyle{\mathop{\sum}_{x^t_d,y^t_d \in \widehat{D^t_d}}} l(\theta,x^t_d,y^t_d).
\end{array}
\end{equation}
\endgroup
% \theta_{t+i-1} - \mathbf{lr_2} \frac{1}{N}\displaystyle{\mathop{\sum}_{j=1}^N}\frac{\partial l(\theta_{t+i-1}, x^i_j,y^i_j)}{\partial \theta}
In these equations, $N$ denotes the size of a batch; $\mathbf{lr}_{1}$ is the learning rate of the sampling distribution; $l(\theta,x,y)$ is the loss of the NMT model on sample $(x,y)$; $J^t(\theta,\vlambda^t)$ is the weighted loss aggregated over $n_d$ dev-sets corresponding to the $n_d$ domains.

@@

\subsection{MDAC for (multi) domain adaptation}
The setting developed in previous sections is quite general and can, in principle, accommodate the variety of situations mentioned above, and many more: basic domain adaption, multi-domain adaptation with various target distributions, possibly including domains unseen in training. In our experiments, we would like to better assess the true potential of MDAC in these settings and seek to experimentally answer the following questions:
\begin{itemize}
\item is MDAC a viable alternative to conventional fine-tuning? In particular, does it enable to better take advantage of relevant data from other domains?
\item is MDAC also a viable option in multi-domain adaptation scenarios?
\item does MDAC also enable to perform \emph{unsupervised} (multi-)domain adaptation? \fyDone{TBContinued}
\end{itemize}
These questions are further explored in section~\ref{sec:results}. We now turn to our experimental conditions.

\section{Experimental settings} \label{sec:exp}
\subsection{Data and metrics \label{ssec:corpora}}
We experiment with translation from English into French and use texts initially originating from 6~domains, corresponding to the following data sources: the UFAL Medical corpus V1.0 (\domain{med})\footnote{\url{https://ufal.mff.cuni.cz/ufal_medical_corpus}. We only use the in-domain (medical) subcorpora: PATR, EMEA, CESTA, ECDC.}; the European Central Bank corpus (\domain{bank}) \citep{Tiedemann12parallel}; the JRC-Acquis Communautaire corpus (\domain{law}) \citep{Steinberger06acquis}; documentations for KDE, Ubuntu, GNOME and PHP from the Opus collection \citep{Tiedemann09news}, merged in a \domain{it}-domain; TedTalks (\domain{talk}) \citep{Cettolo12wit}, and the Koran (\domain{rel}). Complementary experiments also use v12 of the News Commentary corpus (\domain{news}). Most corpora are available from the Opus website.\footnote{\url{http://opus.nlpl.eu}} These corpora were deduplicated and tokenized with in-house tools; statistics are in Table~\ref{tab:Corpora}. To reduce the number of types and build open-vocabulary systems, we use Byte-Pair Encoding \citep{Sennrich16BPE} with 30,000 merge operations on a corpus containing all sentences in both languages.\fyDone{Add \# number of tokens, also specificity ?}%
%
We randomly select in each corpus a development and a test set of 1,000 lines and keep the rest for training. Validation sets are used to chose the best model according to the average BLEU score \citep{Papineni02bleu}.\footnote{We use truecasing and \texttt{sacrebleu}\citep{Post18call}.}\fyDone{A word about meta-parameter settings} Significance testing is performed using bootstrap resampling \citep{Koehn04statistical}, implemented in compare-mt\footnote{\url{https://github.com/neulab/compare-mt}} \citep{Neubig19compare-mt}. We report significant differences at the level of $p=0.05$.\fyDone{Fix correct p value}

@@

\begin{itemize}
\itemsep0em 
\item Generic models trained with predefined mixtures of the training data taking the form:
\begin{align} \label{mixture:trn}
\lambda_{\alpha}(d) = \frac{q_d^{\alpha}}{\displaystyle{\mathop{\sum}_{d=1}^{n_d}q_d^{\alpha}}} &&
q_d = \frac{\mid N^{s}_d \mid}{\displaystyle{N^{s}}} % \mathop{\sum}_{i=1}^K\mid D_i \mid}}
\end{align}
%\setlength{\abovedisplayskip}{2pt}
% \setlength{\belowdisplayskip}{2pt} 
with $\alpha \in \{0,0.25,0.5,0.75,1.0\}$. We denote these as \system{Mixed-$\alpha$} below. \system{Mixed-$0$} uses a uniform distribution, \system{Mixed-$1.0$} the empirical domain distribution.
\item fine-tuned models \citep{Luong15stanford,Freitag16fast}, based on the \system{Mixed-$1.0$} system, further trained on each domain for at most 50~000 iterations, with early stopping when the dev BLEU stops increasing for 5 successive iterations.\fyDone{for ?? iterations} The full fine-tuning (\system{FT-Full}) procedure may update all the parameters of the initial generic model, resulting in six systems, each adapted for one domain, with no parameter sharing across domains.
\item systems trained with fixed data mixtures corresponding to $\vlambda^l \in \big[ \vlambda_0, \vlambda_{0.25}, \vlambda_{0.5}, \vlambda_{0.75}, \vlambda_{1.0}\big]$; these are used in the multi-domain experiments of Section~\ref{ssec:mda};
\item  our own implementations of recent dynamic sampling proposals from the literature: Curriculum Learning (CL) of \citet{Zhang19curriculum} and Differential Data Selection (DDS) of \citet{Wang20balancing} (see details below);
\end{itemize}

@@

In our work, we parameterize the distribution $\vlambda^l$ as follows (with $\beta=2$\footnote{This setting corresponds to the \emph{spherical softmax} of \cite{Brebisson16anexploration}.} in all experiments):\fyDone{Explain why}
\begin{equation}
\lambda^l(d;\vpsi) = \frac{\vpsi[d]^\beta}{\sum_i \vpsi[i]^\beta}. \nonumber
\end{equation}
This parameterization avoids the ``rich-get-richer'' effect that we observe when using $\vlambda(\vpsi) = \operatorname{softmax}(\vpsi)$, which yields gradients wrt. $\psi[d]$ that are proportional to $\exp(\psi[d])$ (see also Figure~\ref{fig:sampling}). Additional settings for the hyper-parameters of our method include the number of simulation steps $k=10$ and the learning rate $\mathbf{lr}_{data}=0.001$. We update the sampling distribution via 100 gradient descent iterations for almost all experimental settings except that for adaptation with automatic clusters (Section~\ref{ssec:clda}), where we use 20~gradient descent iterations to avoid converging to degenerate distributions.\fyFuture{Why? computation?} We split the training into 100 short sessions that last 2000 training steps each. The choice of those hyper-parameters is mostly heuristic except for the learning rate $\mathbf{lr}_{data}$ which is optimized via grid search over a set of values $\{0.001,0.0025,0.005\}$.

@@

\subsection{Experimental tasks}
We evaluate and compare our method with multiple baselines in the 5 following conditions.
%including: domain adaptation; multi-domain adaptation, bi-domain adaptation, unseen domain adaptation, and unsupervised domain adaptation.
In the \emph{supervised domain adaptation task}, given the data from 6 domains (\domain{med}, \domain{bank}, \domain{law}, \domain{it}, \domain{talk}, \domain{rel}), we aim to build distinct expert NMT models for each domain. To challenge the flexibility of the method, we also consider a \emph{bi-domain adaptation task}, where given the same 6 domains, we only focus on adapating only to 2 domains.

@@

A second observation is that DDS is again outperformed by MDAC by a significant margin of 1.5 BLEU on average; the only domain where DDS does (much) better is \domain{med}. Figure~\ref{fig:sampling}, where we plot the evolution of the training mixture in the course of training sessions, helps understand the difference between the two methods. For DDS (Figure~\ref{fig:DDS}), the sampling distribution quickly converges towards a bi-domain mode in which only \domain{med} and \domain{rel} have significant probability -- hence the good performance on the former domain. In contrast, the distribution computed by MDAC evolves more smoothly; small domains such as \domain{bank}, \domain{it}, \domain{talk} and \domain{rel} receive a larger proportion of training data in the early stages; their weights then slowly decrease as larger domains such as \domain{med} and \domain{law} increase their share. This, however, only happens at the end of the training, when the NMT models might have already been close to their optimal performance for the small domains.

@@

\subsection{Unseen domain}\label{ssec:uda}
The left part of Table~\ref{tab:unsupervised-da} displays the performance on the unseen domain \domain{news} for NMT systems trained with the mixtures $\vlambda^l \in \big[ \vlambda_0, \vlambda_{0.25}, \vlambda_{0.5}, \vlambda_{0.75}, \vlambda_{1.0}\big]$ and with dynamic data selection (MDAC and DDS). These systems have insignificant differences in BLEU, showing that dynamic mixture does not improve the robustness of the NMT system against unseen domains. However, the fact that the performance of MDAC and DDS is close to the best performance is a sign that they can also apply in such challenging situations. % the robustness of the NMT model.

\subsection{Automatic clustering}\label{ssec:clda}
The right part of Table~\ref{tab:unsupervised-da} reports the performance of NMT systems adapted to each domain.\footnote{A more detailed analysis is in the supplementary material.} In comparison to Section~\ref{ssec:da}, the training data is distributed in 30 automatic clusters instead of the 6 original domains. Splitting the train data into small groups provides the learner with extra degrees of freedom when selecting the best distribution. However, as these clusters are built automatically, they are noisier in nature. According to the results in Table~\ref{tab:unsupervised-da}, this scenario is hard both for DDS and MDAC, which perform much worse than for the supervised DA setting. This again signals the importance of initialization: analyzing the clustering, we find that the data for \domain{rel} mostly correspond to one single cluster. When using a uniform initialization, this cluster starts with a very small weight and never succeeds in yielding the (good) performance observed in the DA setting.

@@

\citet{Kumar19reinforcement} use reinforcement techniques (deep Q-learning) to learn the curriculum strategy: in this work, complexity corresponds to difficulty levels which are binned using contrastive data selection. The reward is based on the increase of the development set loss that results from the actual data selection strategy.\fyDone{Alert: what do we do during warm-up ?} The same technique is recently applied to multilingual NMT in \citep{Kumar21learning}. \citet{Zhou20uncertainty} propose another curriculum-based approach which instead relies on \emph{instance uncertainty} as a measure of their difficulty and presents the data sample starting with the easiest (more predictable) first. Another contribution of this paper is an alternative criterium for stopping the training. More related to our problems, \citet{Wang20learning-multi} adapt CL for multi-domain adaptation, where an optimal instance weighting scheme is found using Bayesian optimization techniques. Each step consists of (a) weighting instances based on relevance features, (b) fine-tuning a pretrained model using the weighted training set, and is applied iteratively to train a sequence of models. The one that maximizes the development set performance is finally retained.

\section{Conclusion and outlook \label{sec:conclusion}}
\fyDone{Write conclusion}
In this study, we have presented a generic framework to perform a variety of standard adaptation tasks for machine translation, ranging from the conventional supervised domain adaptation to multi-domain adaptation and unseen domain adaptation. By experimenting with all these settings, we have shown that the same algorithm, aimed at automatically finding an effective data sampling scheme during the course of training, could be used in all these situations. This algorithm, we believe, provides us with a more sound approach to (multi-domain) DA than existing heuristics and dispenses with the costly search of optimal meta-parameters. Another contribution of our work is an experimental comparison of recent approaches to dynamic data selection.

>>>>

\begin{document}
\maketitle
\setlength{\abovedisplayskip}{2pt}
\setlength{\belowdisplayskip}{2pt}
\begin{abstract}
  Building effective Neural Machine Translation models often implies accommodating diverse sets of heterogeneous data so as to optimize performance for the domain(s) of interest. Such multi-source / multi-domain adaptation problems are typically approached through instance selection or reweighting strategies, based on a static assessment of the relevance of training instances with respect to the task at hand. In this paper, we study dynamic data selection strategies that are able to automatically re-evaluate the usefulness of data samples and to evolve a data selection policy in the course of training. Based on the results of multiple experiments, we show that such methods constitute a generic framework to automatically and effectively handle a variety of real-world situations, from multi-source domain adaptation to multi-domain learning and unsupervised domain adaptation. 
\end{abstract}

@@

It is usual in MT to opportunistically collect training samples from several domains, which means that training instances are distributed according to a mixture $\mathcal{D}^s$ such that $\mathcal{D}^s(x) = \sum_{d=1}^{n_d} \lambda^{s}(d) \mathcal{D}_d(x)$, with $\{\lambda^{s}(d), d=1 \dots n_d\}$ the mixture weights satisfying $\sum_d \lambda^{s}(d)=1$. In the sequel, we also use boldface to denote the vector $\vlambda$, and $\lambda(d)$ is used to denote a scalar corresponding to the $d^{th}$ element of the vector $\vlambda$.

The main challenge in this situation is to make the best of these heterogeneous data, with the aim to achieve the optimal performance for the intended test conditions. These might correspond to data from just one of the training domains, as in standard supervised (multi-source) domain adaptation; a more difficult case is when the test data is from one domain unseen in training (unseen domain adaptation); in multi-domain adaptation finally, the test distribution is itself a mixture of domains, some of which may also be observed in training.  Without loss of generality, one may then assume that the test distribution takes the form $\mathcal{D}^{t}(x) = \sum_d \lambda^{t}(d) \mathcal{D}_d(x)$ - with only one non-null  component in the case of pure domain adaptation (see illustration on Figure~\ref{fig:mdmt-lambdas}).
\begin{figure}[h]
  \centering
  \includegraphics[width=0.48\textwidth]{mdmt-lambdas}
  \caption{Training and testing with distribution mismatch. We consider just three domains, and represent vectors of mixture weights $\vlambda^{s}$ and $\vlambda^{t}$ in the 3-dimensional simplex. Training with weights in (a) and testing with weights in (c) is supervised multi-source domain adaptation to domain~2 ($d_2$), while (b)-(c) is the unsupervised\fyTodo{Unseen ?} version, with no training data from $d_2$; training with weights in (a) and testing with weights in (d) is multi-domain learning, also illustrated with configurations (a)-(e) (training domain $d_1$ is not seen in test), and (b)-(d)  (test domain $d_2$ is unseen in training).}\label{fig:mdmt-lambdas}
\end{figure}

@@

A widely used approach to supervised DA is \emph{fine-tuning} \citep{Luong15stanford,Freitag16fast}, where $\vlambda^{l}$ is allowed to vary during learning. With our notations, this approach amounts to first learning an initial parameter value with all the data ($\forall d, \lambda^{l}(d) = \lambda^{s}(d)$), then to continue training with only batches from the test domain $d_t$ ($\lambda^{l}(d) = \indic{d = d_t}$) with $\indic{A}$ the indicator function for predicate $A$. Note that this strategy is potentially suboptimal, as some out-of-domain samples may contribute to the final performance due to eg.\ domain similarity. Optimizing the learning distribution in multi-domain settings is even more challenging, as the learner has to best take advantage of potential domains overlaps, and also of the fact that some domains might be easier to learn than others.\fyDone{How to measure this?}
% \revision{Answer: The graphic of \system{Mixed-0} shows that with the same amount of data(batches) the convergences of the domains vary significantly}

@@

% Our method
MDAC constructs an adaptative training distribution $\vlambda^{l}$ that optimizes the data selection policy along with the training of the NMT model. We parameterize $\vlambda^{l}$ by a differentiable function $\vlambda^l(\psi)$, which will be described in detail in Section \ref{ssec:dds-sys}. We divide the training into many short sessions; in each session $t$, the model is trained with a static data distribution $\vlambda^{l}(\psi_t)$. After one learning session, we update the data distribution using the REINFORCE algorithm of \citet{Williams92simple}. The evolution of $\psi$ is thus defined by:
\begin{align*}
\psi_{t+1} &= \psi_t + \mathbf{lr}_{1} * \displaystyle{\mathop{\sum}_{d=1}^{n_d}} R(d) * \frac{\partial \lambda^l(d;\psi_t)}{\partial \psi},
\end{align*}
\begingroup
\allowdisplaybreaks
where the reward $R(d)$ is computed as:
\begin{align*}
  R(d) = J^t(\theta_{t+k},\vlambda^t) - J^t(\theta_t,\vlambda^t),
\end{align*}
and where we also define:
\begin{equation}
\begin{array}{rcl}
\theta_{t+i} &=& Update\big(\theta_{t+i-1},[x^i_j,y^i_j]_{j=1}^N\big) \\ \nonumber
x^i_j, y^i_j &\sim& \widehat{D^l_d}(x) \\
J^t(\theta,\vlambda^t) &=& \displaystyle{\mathop{\sum}_{d=1}^{n_d}}\lambda^t(d)\displaystyle{\mathop{\sum}_{x^t_d,y^t_d \in \widehat{D^t_d}}} l(\theta,x^t_d,y^t_d).
\end{array}
\end{equation}
\endgroup
% \theta_{t+i-1} - \mathbf{lr_2} \frac{1}{N}\displaystyle{\mathop{\sum}_{j=1}^N}\frac{\partial l(\theta_{t+i-1}, x^i_j,y^i_j)}{\partial \theta}
In these equations, $N$ denotes the size of a batch; $\mathbf{lr}_{1}$ is the learning rate of the sampling distribution; $l(\theta,x,y)$ is the loss of the NMT model on sample $(x,y)$; $J^t(\theta,\vlambda^t)$ is the weighted loss aggregated over $n_d$ dev-sets corresponding to the $n_d$ domains.

@@

\subsection{MDAC for (multi) domain adaptation}
The setting developed in previous sections is quite general and can, in principle, accommodate the variety of situations mentioned above, and many more: basic domain adaption, multi-domain adaptation with various target distributions, possibly including domains unseen in training. In our experiments, we would like to better assess the true potential of MDAC in these settings and seek to experimentally answer the following questions:
\begin{itemize}
\item is MDAC a viable alternative to conventional fine-tuning? In particular, does it enable to better take advantage of relevant data from other domains?
\item is MDAC also a viable option in multi-domain adaptation scenarios?
\item does MDAC also enable to perform \emph{unsupervised} (multi-)domain adaptation? \fyDone{TBContinued}
\end{itemize}
These questions are further explored in Section~\ref{sec:results}. We now turn to our experimental conditions.

\section{Experimental settings} \label{sec:exp}
\subsection{Data and metrics \label{ssec:corpora}}
We experiment with translation from English into French and use texts initially originating from 6~domains, corresponding to the following data sources: the UFAL Medical corpus V1.0 (\domain{med})\footnote{\url{https://ufal.mff.cuni.cz/ufal_medical_corpus}. We only use the in-domain (medical) subcorpora: PATR, EMEA, CESTA, ECDC.}; the European Central Bank corpus (\domain{bank}) \citep{Tiedemann12parallel}; the JRC-Acquis Communautaire corpus (\domain{law}) \citep{Steinberger06acquis}; documentations for KDE, Ubuntu, GNOME and PHP from the Opus collection \citep{Tiedemann09news}, merged in a \domain{it}-domain; TedTalks (\domain{talk}) \citep{Cettolo12wit}, and the Koran (\domain{rel}). Complementary experiments also use v12 of the News Commentary corpus (\domain{news}). Most corpora are available from the Opus website.\footnote{\url{http://opus.nlpl.eu}} These corpora were deduplicated and tokenized with in-house tools; statistics are in Table~\ref{tab:Corpora}. To reduce the number of types and build open-vocabulary systems, we use Byte-Pair Encoding \citep{Sennrich16BPE} with 30,000 merge operations on a corpus containing all sentences in both languages.\fyDone{Add \# number of tokens, also specificity ?}%
%
We randomly select in each corpus a development and a test set of 1,000 lines and keep the rest for training. Validation sets are used to chose the best model according to the average BLEU score \citep{Papineni02bleu}.\footnote{We use truecasing and \texttt{sacrebleu} \citep{Post18call}.}\fyDone{A word about meta-parameter settings} Significance testing is performed using bootstrap resampling \citep{Koehn04statistical}, implemented in compare-mt\footnote{\url{https://github.com/neulab/compare-mt}} \citep{Neubig19compare-mt}. We report significant differences at the level of $p=0.05$.\fyDone{Fix correct p value}

@@

\begin{itemize}
\itemsep0em 
\item Generic models trained with predefined mixtures of the training data taking the form:
\begin{align} \label{mixture:trn}
\lambda_{\alpha}(d) = \frac{q_d^{\alpha}}{\displaystyle{\mathop{\sum}_{d=1}^{n_d}q_d^{\alpha}}} &&
q_d = \frac{\mid N^{s}_d \mid}{\displaystyle{N^{s}}} % \mathop{\sum}_{i=1}^K\mid D_i \mid}}
\end{align}
%\setlength{\abovedisplayskip}{2pt}
% \setlength{\belowdisplayskip}{2pt} 
with $\alpha \in \{0,0.25,0.5,0.75,1.0\}$. We denote these as \system{Mixed-$\alpha$} below. \system{Mixed-$0$} uses a uniform distribution, \system{Mixed-$1.0$} the empirical distribution of domains.
\item fine-tuned models \citep{Luong15stanford,Freitag16fast}, based on the \system{Mixed-$1.0$} system, further trained on each domain for at most 50~000 iterations, with early stopping when the dev BLEU stops increasing for 5 successive iterations.\fyDone{for ?? iterations} The full fine-tuning (\system{FT-Full}) procedure may update all the parameters of the initial generic model, resulting in six systems, each adapted for one domain, with no parameter sharing across domains.
\item systems trained with fixed data mixtures corresponding to $\vlambda^l \in \big[ \vlambda_0, \vlambda_{0.25}, \vlambda_{0.5}, \vlambda_{0.75}, \vlambda_{1.0}\big]$; these are used in the multi-domain experiments of Section~\ref{ssec:mda};
\item  our own implementations of recent dynamic sampling proposals from the literature: Curriculum Learning (CL) of \citet{Zhang19curriculum} and Differential Data Selection (DDS) of \citet{Wang20balancing} (see details below);
\end{itemize}

@@

In our work, we parameterize the distribution $\vlambda^l$ as follows (with $\beta=2$ in all experiments):\footnote{This setting corresponds to the \emph{spherical softmax} of \cite{Brebisson16anexploration}.}\fyDone{Explain why}
\begin{equation}
\lambda^l(d;\vpsi) = \frac{\vpsi[d]^\beta}{\sum_i \vpsi[i]^\beta}. \nonumber
\end{equation}
This parameterization avoids the ``rich-get-richer'' effect that we observe when using $\vlambda(\vpsi) = \operatorname{softmax}(\vpsi)$, which yields gradients wrt. $\psi[d]$ that are proportional to $\exp(\psi[d])$ (see also Figure~\ref{fig:sampling}). Additional settings for the hyper-parameters of our method include the number of simulation steps $k=10$ and the learning rate $\mathbf{lr}_{data}=0.001$. We update the sampling distribution via 100 gradient descent iterations for almost all experimental settings except that for adaptation with automatic clusters (Section~\ref{ssec:clda}), where we use 20~gradient descent iterations to avoid converging to degenerate distributions.\fyFuture{Why? computation?} We split the training into 100 short sessions that last 2000 training steps each. The choice of those hyper-parameters is mostly heuristic except for the learning rate $\mathbf{lr}_{data}$ which is optimized via grid search over a set of values $\{0.001,0.0025,0.005\}$.

@@

\subsection{Experimental tasks}
We evaluate and compare our method with multiple baselines in the 5 following conditions.
%including: domain adaptation; multi-domain adaptation, bi-domain adaptation, unseen domain adaptation, and unsupervised domain adaptation.
In the \emph{supervised domain adaptation task}, given the data from 6 domains (\domain{med}, \domain{bank}, \domain{law}, \domain{it}, \domain{talk}, \domain{rel}), we aim to build distinct expert NMT models for each domain. To challenge the flexibility of the method, we also consider a \emph{bi-domain adaptation task}, where given the same 6 domains, we only focus on adapting to a mixture of 2~domains.

@@

A second observation is that MDAC again outperforms DDS by a wide margin (+1.5 BLEU on average); the only domain where DDS does better is \domain{med}. Figure~\ref{fig:sampling}, which displays the evolution of the training mixture in the course of training, helps to understand the difference between the two methods. For DDS (Figure~\ref{fig:DDS}), the sampling distribution quickly reaches a bi-domain mode in which only \domain{med} and \domain{rel} have significant probability -- hence the good performance on the former domain. In contrast, the distribution computed by MDAC evolves more smoothly; small domains such as \domain{bank}, \domain{it}, \domain{talk} and \domain{rel} receive a larger proportion of training data in the early stages; their weights then slowly decrease as larger domains such as \domain{med} and \domain{law} increase their share. This, however, only happens at the end of training, when some NMT models might already be close to their best performance for the small domains.

@@

\subsection{Unseen domain}\label{ssec:uda}
The left part of Table~\ref{tab:unsupervised-da} displays the performance on the unseen domain \domain{news} for NMT systems trained with the mixtures $\vlambda^l \in \big[ \vlambda_0, \vlambda_{0.25}, \vlambda_{0.5}, \vlambda_{0.75}, \vlambda_{1.0}\big]$ and with dynamic data selection (MDAC and DDS). These systems have insignificant differences in BLEU, showing that dynamic mixture does not improve the robustness of the NMT system against unseen domains. However, the fact that the performance of MDAC and DDS is close to the best performance suggests that they also apply in such challenging situations. % the robustness of the NMT model.

\subsection{Automatic clustering}\label{ssec:clda}
The right part of Table~\ref{tab:unsupervised-da} reports the performance of NMT systems adapted to each domain.\footnote{A more detailed analysis is in the supplementary material.} In comparison to Section~\ref{ssec:da}, the training data is distributed in 30 automatic clusters instead of the 6 original domains. Splitting the train data into small groups provides the learner with extra degrees of freedom when selecting the best distribution. However, as these clusters are built automatically, they are noisier in nature. According to the results in Table~\ref{tab:unsupervised-da}, this scenario is hard both for DDS and MDAC, which performs much worse than for the supervised DA setting. This again signals the importance of initialization: analyzing the clustering, we find that the data for \domain{rel} mostly correspond to one single cluster. When using a uniform initialization, this cluster starts with a very small weight and never succeeds in matching the (good) performance observed in the DA setting.

@@

\citet{Kumar19reinforcement} use reinforcement techniques (deep Q-learning) to learn the curriculum strategy: in this work, complexity corresponds to difficulty levels which are binned using contrastive data selection. The reward is based on the increase of the development set loss that results from the actual data selection strategy.\fyDone{Alert: what do we do during warm-up ?} The same technique is recently applied to multilingual NMT in \citep{Kumar21learning}. \citet{Zhou20uncertainty} propose another curriculum-based approach which instead relies on \emph{instance uncertainty} as a measure of their difficulty and presents the data sample starting with the easiest (more predictable) first. Another contribution of this paper is an alternative stopping criterium. Closest to our problems, \citet{Wang20learning-multi} adapt CL for multi-domain adaptation, where an optimal instance weighting scheme is found using Bayesian optimization techniques. Each step consists of (a) weighting instances based on relevance features, (b) fine-tuning a pretrained model using the weighted training set, and is applied iteratively to train a sequence of models. The one that maximizes the development set performance is finally retained.

\section{Conclusion and outlook \label{sec:conclusion}}
\fyDone{Write conclusion}
In this study, we have presented a generic framework to perform a variety of adaptation tasks for machine translation, ranging from conventional supervised domain adaptation to multi-domain adaptation and unseen domain adaptation. By experimenting with all these settings, we have shown that the same algorithm, aimed at automatically finding an effective data sampling scheme during the course of training, could be used in all these situations. This algorithm, we believe, provides us with a more sound approach to (multi-domain) DA than existing heuristics and dispenses with the costly search of optimal meta-parameters. Another contribution of our work is an experimental comparison of recent approaches to dynamic data selection.
